Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-06-07 paper/intro.tex
Lots of small edits + some new XMT results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@628 3fc257bb-8550-0410-81b3-a5c103da83f7

Similarly, the open-source Hadoop implementation of MapReduce~\cite{Hadoop}, 
has become widely popular in the past few years for
parallel analysis of large-scale data sets at Yahoo and other
data-centric companies, as well as in university and laboratory
research groups, due to its free availability.  MapReduce programs in
Hadoop are typically written in Java, though it also supports use of
stand-alone map and reduce kernels, which can be written as shell
scripts or in other languages.

@@

\item {\it C++ library using MPI for inter-processor communication:}
The user writes a (typically) simple main program which runs on each
processor of a parallel machine, making calls to the MR-MPI library.
For map and reduce operations, the library calls back to user-provided
map() and reduce() functions.  The use of C++ allows precise control
over the memory and format of data allocated by each processor during
a MapReduce.  Library calls for performing a map, reduce, or data
shuffle, are synchronous, meaning all the processors participate and
finish the operation before proceeding.  Similarly, the use of MPI
within the library is the traditional mode of MPI\_Send and
MPI\_Recv calls between processor pairs using large aggregated
messages to improve bandwidth performance and reduce latency costs.  A
recent paper~\cite{Hoefler09} also outlines the MapReduce formalism
from an MPI perspective, though they advocate a more asyncrhonous
approach, using one-way communication of small messages.

\item {Small, portable:} The entire MR-MPI library is a few thousand
lines of standard C++ code.  For parallel operation, the program is
linked with MPI, a standard message passing library available on all
distributed memory machines.  For serial operation, a dummy MPI
library (provided) can be substituted.  As a library, it can be
embedded in other codes~\cite{Titan} to enable them to
perform MapReduces.

\item {\it In-core or out-of-core operation:} Each MapReduce object that a
processor defines allocates per-processor ``pages'' of memory, where
the page size is determined by the user.  Typical MapReduce operations
can be performed using a few such pages.  If the data set fits in a
single page (per processor), then the library performs its operations
in-core.  If the data set exceeds the page size, then processors each
write to temporary disk files (to local disk or a parallel file
system) as needed and subsequently read from them.  This allows
processing of data sets larger than the aggregate memory of all the
processors, i.e. up to the available aggregate disk space.

\item {\it Flexible programmability:} An advantage of writing a
MapReduce program on top of MPI, is that the user program can invoke
MPI calls directly, if desired.  For example, one-line calls to
MPI\_Allreduce are often useful in determining the status of an
iterative graph algorithm, as described in Section~\ref{sec:graph}.
The library interface also provides a user data pointer as an argument
passed back to all callback functions, so it is easy for the user
program to store ``state'' on each processor, accessible during the
map and reduce operations.  For example, various flags can be stored
that alter the operation of a map or reduce function, as can richer
data structures, that accumulate the results.

\item {\it C++, C, and Python interfaces:} A C++ interface to the
MR-MPI library means a user program instantiates and then invokes
methods in one or more MapReduce objects.  A C interface means the
library can also be called from C or other high-level languages such as
Fortran.  A C interface also means the library can be easily wrapped
by Python via the Python ``ctypes'' module.  The library can then be
called from a Python script, allowing the user to write map() and
reduce() callback functions in Python.  If a machine supports running
Python in parallel, a parallel MapReduce can also be run in this mode.

\item {No fault tolerance:} Current MPI implementations do not enable
easy detection of a dead processor, or retrieval of the data it was
working on.  So like most MPI programs, a parallel program calling the
MR-MPI library will hang or crash if a processor goes away.  Unlike
Hadoop, and its HDFS file system which provides for data redundancy,
the MR-MPI library simply reads and writes simple, flat files.  It can
use local per-processor disks, or a parallel file system, if
available, but these typically provide no data redundancy.

@@

The remainder of the paper is organized as follows.  The next two
sections~\ref{sec:mr} and~\ref{sec:outcore} describe how in-core and
out-of-core MapReduce primitives are formulated as MPI-based
operations in the MR-MPI library.  Section~\ref{sec:graph} briefly
describes the formulation of several common graph algorithms as
MapReduce operations.  Section~\ref{sec:results} gives performance
results for these algorithms running on a parallel cluster for graphs
%KDDKDD CHECK SIZES
ranging in size from 1 million to 2 billion vertices or edges.  In
this section, we highlight the performance and complexity trade-offs
of a MapReduce approach versus other more special-purpose algorithms.
The latter generally perform better but are harder to implement
efficiently on distributed memory machines, due to the required
explicit management of parallelism, particularly for large out-of-core
data sets.  Section~\ref{sec:lessons} summarizes some lessons learned
from the implementation and use of our library.

>>>>

Similarly, the open-source Hadoop implementation of MapReduce~\cite{Hadoop}, 
has become widely popular in the past few years for
parallel analysis of large-scale data sets at Yahoo and other
data-centric companies, as well as in university and laboratory
research groups, due to its free availability.  MapReduce programs in
Hadoop are typically written in Java, although it also supports use of
stand-alone map and reduce kernels, which can be written as shell
scripts or in other languages.

@@

\item {\it C++ library using MPI for inter-processor communication:}
The user writes a (typically) simple main program which runs on each
processor of a parallel machine, making calls to the MR-MPI library.
For map and reduce operations, the library calls back to user-provided
{\it map()} and {\it reduce()} functions.  The use of C++ allows precise control
over the memory and format of data allocated by each processor during
a MapReduce.  Library calls for performing a map, reduce, or data
shuffle, are synchronous, meaning all the processors participate and
finish the operation before proceeding.  Similarly, the use of MPI
within the library is the traditional mode of MPI\_Send and
MPI\_Recv calls between processor pairs using large aggregated
messages to improve bandwidth performance and reduce latency costs.  A
recent paper~\cite{Hoefler09} also outlines the MapReduce formalism
from an MPI perspective, although they advocate a more asynchronous
approach, using one-way communication of small messages.

\item {\it Small, portable:} The entire MR-MPI library is a few thousand
lines of standard C++ code.  For parallel operation, the program is
linked with MPI, a standard message passing library available on all
distributed memory machines.  For serial operation, a dummy MPI
library (provided) can be substituted.  As a library, it can be
embedded in other codes~\cite{Titan} to enable them to
perform MapReduce operations.

\item {\it In-core or out-of-core operation:} Each MapReduce object that a
processor defines allocates per-processor ``pages'' of memory, where
the page size is determined by the user.  Typical MapReduce operations
can be performed using a few such pages.  If the data set fits in a
single page (per processor), the library performs its operations
in-core.  If the data set exceeds the page size, processors each
write to temporary disk files (on local disks or a parallel file
system) as needed and subsequently read from them.  This out-of-core
operation allows
processing of data sets larger than the aggregate memory of all the
processors, i.e. up to the available aggregate disk space.

\item {\it Flexible programmability:} An advantage of writing a
MapReduce program on top of MPI, is that the user program can invoke
MPI calls directly, if desired.  For example, one-line calls to
MPI\_Allreduce are often useful in determining the status of an
iterative graph algorithm, as described in Section~\ref{sec:graph}.
The library interface also provides a user data pointer as an argument
passed back to all callback functions, so it is easy for the user
program to store ``state'' on each processor, accessible during the
map and reduce operations.  For example, various flags can be stored
that alter the operation of a map or reduce operation, as can richer
data structures that accumulate results.

\item {\it C++, C, and Python interfaces:} The C++ interface to the
MR-MPI library allows user programs to instantiate and then invoke
methods in one or more MapReduce objects.  The C interface allows the
library to be called from C or other high-level languages such as
Fortran.  The C interface also allows the library to be wrapped easily
by Python via the Python ``ctypes'' module.  The library can then be
called from a Python script, allowing the user to write {\it map()} and
{\it reduce()} callback functions in Python.  If a machine supports running
Python in parallel, a parallel MapReduce can also be run in this mode.

\item {\it No fault tolerance:} Current MPI implementations do not enable
easy detection of a dead processor or retrieval of the data it was
working on.  So like most MPI programs, a parallel program calling the
MR-MPI library will hang or crash if a processor goes away.  Unlike
Hadoop, and its HDFS file system which provides data redundancy,
the MR-MPI library reads and writes simple, flat files.  It can
use local per-processor disks, or a parallel file system, if
available, but these typically provide no data redundancy.

@@

The remainder of the paper is organized as follows.  
Sections~\ref{sec:mr} and~\ref{sec:outcore} describe how in-core and
out-of-core MapReduce primitives are formulated as MPI-based
operations in the MR-MPI library.  Section~\ref{sec:graph} briefly
describes the formulation of several common graph algorithms as
MapReduce operations.  Section~\ref{sec:results} gives performance
results for these algorithms running on a parallel cluster for graphs
%KDDKDD CHECK SIZES
ranging in size from 8 million to 2 billion edges;  
we highlight the performance and complexity trade-offs
of a MapReduce approach versus other more special-purpose algorithms.
The latter generally perform better but are harder to implement
efficiently on distributed memory machines, due to the required
explicit management of parallelism, particularly for large out-of-core
data sets.  Section~\ref{sec:lessons} summarizes some lessons learned
from the implementation and use of our library.

### 2010-06-07 paper/lessons.tex
Lots of small edits + some new XMT results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@628 3fc257bb-8550-0410-81b3-a5c103da83f7

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a
shuffle) with communication between all pairs of processors.  But the
benefit is often good load-balance, even for hard-to-balance irregular
data sets.  By contrast, more traditional distributed-memory parallel
algorithms, e.g. for matrix operations, or grid-based or
particle-based simulation codes, tend to work hard to localize data
and minimize communication.  To do this they typically require a lot
of application-specific logic and parallel communication coding, to
create and maintain a data decomposition, generate ghost versions of
nearby spatially-decomposed data, etc.

MapReduce algorithms can be hard to design, but are often relatively
easy to write and debug.  Thinking about a computational task from a
MapReduce perspective is different from traditional distributed-memory
parallel algorithm design.  For example, with an MPI mindset, it often
seems heretical to intentionally ignore data locality.  However,
writing small {\it map()} and {\it reduce()} functions is typically easy.  And
writing an algorithm that involves complex parallel operations,
without actually needing to write application-specific parallel code
to move data via MPI calls, is often a pleasant surprise.  Moreover,
if the MapReduce algorithm is initially coded so that it runs
correctly on one processor, it often works out-of-the-box on hundreds
or thousands of processors, without the need for additional debugging.

Performing MapReduce operations on a fixed allocation of processors on
a traditional MPI-based parallel machine is a somewhat different
conceptual model than that of cloud-computing MapReduces using (for
example) Hadoop.  In the former case, one can potentially control
which processor owns which data at various stages of an algorithm.
This is somewhat hidden from the user in a typical cloud-computing
model, where data simply exists somewhere in the cloud and Hadoop
ensures data moves where it needs to and is operated on by some
processor.  The cloud model is a nice data-centric abstraction which
allows for fault tolerance both to data loss (via redundant storage)
and to processor failure (via reassignment of work), neither of which
is typically possible on current MPI-based parallel machines.

However, since the MPI implementation of MapReduce, at least as
described in this paper, is processor-centric, one can sometimes
fruitfully exploit the possibility for processors to maintain
``state'' over the course of multiple map and reduce operations.  By
controlling where data resides for maps and reduces (e.g. via a
user-specified hash function), and by assuming that processor will
always be available, more efficient operations with less data movement
are sometimes possible.  The discussion of enhanced graph algorithms
in Section \ref{sec:graph} illustrated this.  To fully exploit this
idea for large data sets, mechanisms are needed for processors to
store, retrieve, and efficiently find needed datums on local disk
(e.g. static edges of a graph), so that archived state can be used
efficiently during subsequent map and reduce operations.

>>>>

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a
shuffle) with communication between all pairs of processors.  But the
benefit is often good load-balance, even for hard-to-balance irregular
data sets.  By contrast, more traditional distributed-memory parallel
algorithms, e.g. for matrix operations, or grid-based or
particle-based simulation codes, tend to work hard to localize data
and minimize communication.  To do this they typically require a lot
of application-specific logic and parallel communication coding to
create and maintain a data decomposition, generate ghost versions of
nearby spatially-decomposed data, etc.

MapReduce algorithms can be hard to design, but are often relatively
easy to write and debug.  Thinking about a computational task from a
MapReduce perspective is different from traditional distributed-memory
parallel algorithm design.  For example, with an MPI mindset, it often
seems heretical to intentionally ignore data locality.  However,
writing small {\it map()} and {\it reduce()} functions is typically easy.  And
writing an algorithm that involves complex parallel operations
without actually needing to write application-specific parallel code
to move data via MPI calls is often a pleasant surprise.  Moreover,
if the MapReduce algorithm is initially coded so that it runs
correctly on one processor, it often works out-of-the-box on hundreds
or thousands of processors, without the need for additional debugging.

Performing MapReduce operations on a fixed allocation of processors on
a traditional MPI-based parallel machine is a somewhat different
conceptual model than that of cloud-computing MapReduces using (for
example) Hadoop.  In the former case, one can potentially control
which processor owns which data at various stages of an algorithm.
This is somewhat hidden from the user in a typical cloud-computing
model, where data simply exists somewhere in the cloud and Hadoop
ensures data moves where it is needed and is operated on by some
processor.  The cloud model is a nice data-centric abstraction which
allows for fault tolerance both to data loss (via redundant storage)
and to processor failure (via reassignment of work), neither of which
is typically possible on current MPI-based parallel machines.

However, since the MPI implementation of MapReduce as
described in this paper is processor-centric, one can sometimes
fruitfully exploit the possibility for processors to maintain
``state'' over the course of multiple map and reduce operations.  By
controlling where data resides for maps and reduces (e.g. via a
user-specified hash function), and by assuming that processor will
always be available, more efficient operations with less data movement
are sometimes possible.  The discussion of enhanced graph algorithms
in Section \ref{sec:graph} illustrated this.  To fully exploit this
idea for large data sets, mechanisms are needed for processors to
store, retrieve, and efficiently find needed datums on local disk
(e.g. static edges of a graph), so that archived state can be used
efficiently during subsequent map and reduce operations.

### 2010-06-07 paper/mr.tex
Lots of small edits + some new XMT results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@628 3fc257bb-8550-0410-81b3-a5c103da83f7

The basic datums stored and operated on by any MapReduce framweork are
key/value (KV) pairs.  In the MR-MPI library, individual keys or
values can be of any data type or length, or combinations of multiple
types (one integer, a string of characters, two integers and a double,
etc); they are simply treated as byte strings by the library.  A KV
pair always has a key; its value may be NULL.  A related data type is
the key/multivalue (KMV) pair, where all values associated with the
same key are collected and stored contiguously as a multivalue, which
is just a longer byte string with an associated vector of integer
lengths, one per value.

A typical MR-MPI program makes at least three calls to the MR-MPI library,
to perform a {\it map()}, {\it collate()}, and {\it reduce()}
operation.  In a {\it map()}, zero or more key/value pairs are
generated by each processor.  Typically, this is done using data read
in from files, but a {\it map()} may generate data itself or process
existing KV pairs to create new ones.  The KV pairs produced are
stored locally by each processor; a {\it map()} thus requires no
inter-processor communication.  Users call the library with a count of
tasks to perform and a pointer to a user function; the MR-MPI {\it
map()} operation invokes the user function multiple times as a
callback.  Depending on which variant of {\it map()} is called, the
user function may be passed a file name, a chunk of bytes from a large
file, a task ID, or a KV pair.  Options for assigning map tasks to processors
are specified by the user and include assigning consecutive 
chunks of tasks to processors, striding the tasks across processors,
or using a master-slave model that is useful when tasks have widely
varying workloads.

The {\it collate()} operation (or data shuffle in Hadoop) identifies
unique keys and collects all the values associated with those keys to
create KMV pairs.  This is done in two stages, the first of which
requires communication, since KV pairs with the same key may be owned
by any processor.  Each processor hashes each of its keys to determine
which processor will ``own'' it.  The $k$-byte length key is hashed into
a 32-bit value whose remainder modulo $P$ processors generates the owning
processor ID.  Alternatively, the user can provide a hash function
which converts the key into a processor ID.  Each processor then sends
each of its KV pairs to the owning processor.

After receiving new KV pairs, the second stage is an on-processor
computation, requiring no further communication.  Each processor
reorganizes its KV pairs into KMV pairs, one for each unique key it
owns.  This is done using a hash table, rather than a sort.  A sort
requires $\log(N)$ passes through the $N$ KV pairs.  With hashing,
the list of KMV
pairs can be created in two passes through the KV pairs, one to populate
the hash table with needed count and length information, and the
second pass to copy the key and value datums into the appropriate
location in a new KMV data structure.  Since the lookup of a key in a
well-formed hash table is a constant-time $O(1)$ operation, the cost
of the data reorganization is $O(N)$, rather than $O(N\log(N))$ for full
sorting.  It also has the added benefit that the number of values in each
KMV pair is known and is passed to the user function during a
reduce.  For some reduce operations, this count is all the information a
reduce requires; thus the values need not be looped over.  This feature is not
available in Hadoop-style data reorganizations, which sort the values; 
with sorting, the number of values associated with the same key is not known 
without interating over the values.

@@

The interface to various low-level operations is provided so
that a user's program can string them together in various ways to
produce interesting MapReduce algorithms.  For example, output from a
{\it reduce()} can serve as input to a subsequent {\it map()} or {\it
collate()}.  KV pairs from mutiple MapReduce objects can be combined
together to perform new sequences of {\it map()}, {\it collate()}, and
{\it reduce()} operations.

The above discussion assumed that the KV or KMV pairs stored by a
processor fit in its physical memory.  If this is the case, then we
refer to this as ``in-core'' processing and no disk files are written
or read by any of the processors, aside from initial input data if it
exists or final output data if it is generated.  Note that the
aggregate physical memory of large parallel machines can be multiple
terabytes, which allows for large data sets to be processed in-core,
assuming the KV and KMV pairs remain evenly distributed across
processors throughout the sequence of MapReduce operations.  The use
of randomized hashing to assign keys to processors is designed for
such load-balancing.  In the next section, we discuss what happens 
when data sets do not fit in available memory.

>>>>

The basic datums stored and operated on by any MapReduce framweork are
key/value (KV) pairs.  In the MR-MPI library, individual keys or
values can be of any data type or length, or combinations of multiple
types (one integer, a string of characters, two integers and a double,
etc); they are simply treated as byte strings by the library.  A KV
pair always has a key; a KV's value may be NULL.  A related data type is
the key/multivalue (KMV) pair, where all values associated with the
same key are collected and stored contiguously as a multivalue, which
is just a longer byte string with an associated vector of integer
lengths, one per value.

A typical MR-MPI program makes at least three calls to the MR-MPI library,
to perform a {\it map()}, {\it collate()}, and {\it reduce()}
operation.  In a {\it map()}, zero or more key/value pairs are
generated by each processor.  Typically, this is done using data read
from files, but a {\it map()} may generate data itself or process
existing KV pairs to create new ones.  The KV pairs produced are
stored locally by each processor; a {\it map()} thus requires no
inter-processor communication.  Users call the library with a count of
tasks to perform and a pointer to a user function; the MR-MPI {\it
map()} operation invokes the user function multiple times as a
callback.  Depending on which variant of {\it map()} is called, the
user function may be passed a file name, a chunk of bytes from a large
file, a task ID, or a KV pair.  Options for assigning map tasks to processors
are specified by the user and include assigning consecutive 
chunks of tasks to processors, striding the tasks across processors,
or using a master-slave model that is useful when tasks have widely
varying workloads.

The {\it collate()} operation (or data shuffle in Hadoop) identifies
unique keys and collects all the values associated with those keys to
create KMV pairs.  This is done in two stages, the first of which
requires communication, since KV pairs with the same key may be owned
by any processor.  Each processor hashes each of its keys to determine
which processor will ``own'' it.  The $k$-byte length key is hashed into
a 32-bit value whose remainder modulo $P$ processors generates the owning
processor rank.  Alternatively, the user can provide a hash function
which converts the key into a processor rank.  Each processor then sends
each of its KV pairs to the owning processor.

After receiving new KV pairs, the second stage is an on-processor
computation, requiring no further communication.  Each processor
reorganizes its KV pairs into KMV pairs, one for each unique key it
owns.  This is done using a hash table, rather than a sort.  A sort
requires $\log(N)$ passes through the $N$ KV pairs.  With hashing,
the list of KMV
pairs can be created in two passes through the KV pairs. The first 
pass populates
the hash table with needed count and length information; the
second pass copies the key and value datums into the appropriate
location in a new KMV data structure.  Since the lookup of a key in a
well-formed hash table is a constant-time $O(1)$ operation, the cost
of the data reorganization is $O(N)$, rather than $O(N\log(N))$ for full
sorting.  It also has the added benefit that the number of values in each
KMV pair is known and is passed to the user function during a
reduce.  For some reduce operations, this count is all the information a
reduce requires; thus the values need not be looped over.  This feature is not
available in Hadoop-style data reorganizations, which sort the values; 
with sorting, the number of values associated with the same key is not known 
without interating over the values.

@@

The interface to various low-level operations is provided so
that a user's program can string them together in various ways to
produce interesting MapReduce algorithms.  For example, output from a
{\it reduce()} can serve as input to a subsequent {\it map()} or {\it
collate()}.  KV pairs from multiple MapReduce objects can be combined
to perform new sequences of {\it map()}, {\it collate()}, and
{\it reduce()} operations.

The above discussion assumed that the KV or KMV pairs stored by a
processor fit in its physical memory.  In this case, MR-MPI 
performs only ``in-core'' processing and no disk files are written
or read by any of the processors, aside from initial input data if it
exists or final output data if it is generated.  Note that the
aggregate physical memory of large parallel machines can be multiple
terabytes, which allows for large data sets to be processed in-core,
assuming the KV and KMV pairs remain evenly distributed across
processors throughout the sequence of MapReduce operations.  The use
of hashing to assign keys to processors is designed for
such load-balancing.  In the next section, we discuss what happens 
when data sets do not fit in available memory.

### 2010-06-07 paper/outcore.tex
Lots of small edits + some new XMT results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@628 3fc257bb-8550-0410-81b3-a5c103da83f7

We now explain how the MapReduce operations described in the previous
section work in out-of-core mode.  The {\it map()} and {\it reduce()}
operations are relatively simple.  As a {\it map()} generates KV pairs
via the user function, a page of memory fills up, one KV pair at a
time.  When the page is full, it is written to disk.  If the source of
data for the {\it map()} operation is an existing set of KV pairs,
then those datums are read in, one page at a time, and a pointer to
each KV pair is given to the user function.  Similarly, a {\it
reduce()} reads one page of KMV pairs from disk, passes a pointer, one
at a time, to each pair to the user function which typically generates
new KV pairs.  The generated pairs fill up a new page, which is
written to disk when full, just as with the {\it map()} operation.
Thus for both a {\it map()} and {\it reduce()}, out-of-core disk files
are read and written sequentially, one page at a time, which requires
at most two pages of memory.

A special case is when a single KMV pair is larger than a single page.
This can happen, for example, in a connected component finding
algorithm if the graph collapses into one or a few giant components.
In this case, the set of values (graph vertices and edges) associated
with a unique key (the component ID), may not fit in one page of
memory.  The individual values in the multivalue will then be spread
across as many pages as needed.  The user function that processes the
KMV pair is passed a flag indicating it received only a portion of the
values.  Once it has processed them, it can request a new set of
values from the MR-MPI library, which reads in a new page from disk.

Performing an our-of-core {\it collate()} operation is more complex.
Recall that the operation occurs in two stages.  First, keys are
hashed to ``owning'' processors and the KV datums are communicated to
new processors in an all-to-all fashion.  In out-of-core mode, this 
operation is
performed on one page of KV data at a time.  Each processor reads in a page of
KV pairs, the communication pattern (which processors will receive 
which datums in that page) is
determined, and all-to-all communication is
performed.  Each processor allocates a two-page chunk of memory to
receive incoming KV pairs.  On average each processor should receive
one page of KV pairs; the two-page allocation allows for some
load-imbalance.  If the KV pairs are distributed unevenly so that this
limit is exceeded, then the all-to-all communication is performed in
smaller, multiple passes until the full page contributed by every
processor has been communicated.  Performing an MPI\_Alltoall(), or
using the custom all-to-all routines provided by the MR-MPI library,
requires allocation of auxiliary arrays that store processor indices,
datum pointers, and datum lengths.  For the case of many tiny KV
datums, the total number of memory pages required to perform the
all-to-all communication, including the source and target KV pages is
at most seven.

When the communication stage of the {\it collate()} operation is
complete, each processor has a set of KV pages, stored in an
out-of-core KV file, that need to be reorganized into a set of KMV
pages, stored in a new out-of-core KMV file.  In principle, 
creating one page of KMV pairs would require all KV pages be scanned
to find all the keys that contribute values to that page.  Doing this
for each output page of KMV pairs could be prohibitively expensive.  A
related issue is that, as described in the previous section, a hash
table is needed to match new keys with previously encountered keys.
But there is no guarantee that the hash table itself will not grow
arbitrarily large for data sets with many unique keys.  We need an
algorithm for generating KMV pairs that operates within a small
number of memory pages and performs a minimal number of passes through
the KV disk file.  An algorithm that meets these goals is diagrammed
in Figure \ref{fig:collate}; it reads the KV pages at most four times, and
writes out new KV or KMV pages at most three times.  It also uses a
finite-size in-memory hash table, which stores unique keys as they are
encountered while looping over the KV pages, as well as auxiliary
information needed to construct the output pages of KMV pairs.

@@

\item (Pass 1) The KV pairs are read, one page at a time, and split
into ``partition'' files, represented by KVp in the figure.  Each
partition file contains KV pairs whose unique keys will (likely) fit
in the hash table (HT).  Initially, all KV pairs are assigned to the
first partition file as the HT is populated.  When the HT becomes
full, e.g. at the point represented by the horizontal line shown on
the leftmost vertical line as a downward scan is performed, then the
fraction of KV pairs read thus far is used to estimate the number of
additional partition files needed.  As subsequent KV pairs are read
in, they are assigned to the original partition file if the KV pair's
key is already in the current HT.  If not, then a subset of bits in
the hash value of the key is used to assign the KV pair to one of the
new partition files.  The number of needed partition files is
estimated conservatively, rounding up to the next power-of-two, so
that extra passes through the KV pairs are almost never needed, and so
that the bit masking can be done quickly.  This first pass entails
both a read and write of all the KV pairs.

@@

\item (Pass 3) The unique keys in the HT for one partition are
scanned to determine the total size of KMV for keys in the HT.  
The associated values may create KMV pairs that will span
$M$ memory pages.  If $M > 1$ the associated partition file of KV
pairs is read again, and each KV pair is assigned to one of $M$
smaller ``set'' files, represented by KVs in the figure.  Each set
file contains the KV pairs that will contribute to the KMV pairs that
will populate one page of KMV output.  If a single KMV pair spans
multiple pages, because it contains a very large number of values,
then the corresponding KV pairs will still be written to a single set
file.  Eventually, this pass reads all partition files, entailing both
a read and write of all KV pairs.

@@

The memory page requirements for the above algorithm are as follows.
Two contiguous pages of memory are used for the hash table.
Intermediate passes 2 and 3 can require numerous partition or set
files to be opened simultaneously and written to.  To avoid small
writes of individual KV datums, a buffer of at least 16K bytes is
allocated for each file.  The precise number of simultaneously open
files is difficult to bound, but typically one or two additional pages'
memory suffice for all needed buffers.  Thus the total number of pages required
for the data reorganization stage of the {\it collate()} operation is
less than the seven used by the all-to-all communication stage.

Other MR-MPI library calls discussed in the previous section, such as
{\it compress()}, {\it convert()}, {\it collapse()}, and {\it gather()},
can also operate in out-of-core mode, typically with one pass
through the KV or KMV data and the use of one or two memory pages.
Sorting KV pairs, by key or value, is an exception.  An out-of-core
merge sort is performed as follows.  Two pages of KV datums are read
from disk.  Each is sorted using the C-library quicksort() function,
as discussed in the previous section.  The two pages are then scanned
and merged into a new file which is written to disk as its associated
memory page fills up.  This process requires five memory pages, which
includes vectors of KV datum pointers and lengths needed by the
in-memory quicksort() operation.  Once all pairs of pages have been
merged, the sort continues in a recursive fashion, merging pairs of
files into a new third file, without the need to perform additional
in-memory sorts.  Thus the overall memory requirement is five pages.  The
number of read/write passes through the KV data set for the merge sort
is $O(\log{M})$, where $M$ is the number of pages of KV pairs.

>>>>

We now explain how the MapReduce operations described in the previous
section work in out-of-core mode.  The {\it map()} and {\it reduce()}
operations are relatively simple.  As a {\it map()} generates KV pairs
via the user function, a page of memory fills up, one KV pair at a
time.  When the page is full, it is written to disk.  If the source of
data for the {\it map()} operation is an existing set of KV pairs,
those datums are read, one page at a time, and a pointer to
each KV pair is given to the user function.  Similarly, a {\it
reduce()} reads one page of KMV pairs from disk, passes a pointer
to each pair, one at a time, to the user function that typically generates
new KV pairs.  The generated pairs fill up a new page, which is
written to disk when full, just as with the {\it map()} operation.
Thus for both a {\it map()} and {\it reduce()}, out-of-core disk files
are read and written sequentially, one page at a time, which requires
at most two pages of memory.

A special case is when a single KMV pair is larger than a single page.
This can happen, for example, in a connected component finding
algorithm if the graph collapses into one or a few giant components.
In this case, the set of values (graph vertices and edges) associated
with a unique key (the component ID), may not fit in one page of
memory.  The individual values in the multivalue are then spread
across as many pages as needed.  The user function that processes the
KMV pair is passed a flag indicating it received only a portion of the
values.  Once it has processed them, it can request a new set of
values from the MR-MPI library, which reads in a new page from disk.

Performing an our-of-core {\it collate()} operation is more complex.
Recall that the operation occurs in two stages.  First, keys are
hashed to ``owning'' processors and the KV datums are communicated to
new processors in an all-to-all fashion.  In out-of-core mode, this 
operation is
performed on one page of KV data at a time.  Each processor reads in a page of
KV pairs, the communication pattern (which processors receive 
which datums in that page) is
determined, and all-to-all communication is
performed.  Each processor allocates a two-page chunk of memory to
receive incoming KV pairs.  On average each processor should receive
one page of KV pairs; the two-page allocation allows for some
load imbalance.  If the KV pairs are distributed unevenly so that this
limit is exceeded, the all-to-all communication is performed in
smaller, multiple passes until the full page contributed by every
processor has been communicated.  Performing an MPI\_Alltoall(), or
using the custom all-to-all routines provided by the MR-MPI library,
requires allocation of auxiliary arrays that store processor indices,
datum pointers, and datum lengths.  For the case of many tiny KV
datums, the total number of memory pages required to perform the
all-to-all communication, including the source and target KV pages is
at most seven.

When the communication stage of the {\it collate()} operation is
complete, each processor has a set of KV pages, stored in an
out-of-core KV file, that need to be reorganized into a set of KMV
pages, stored in a new out-of-core KMV file.  In principle, 
creating one page of KMV pairs would require all KV pages be scanned
to find all the keys that contribute values to that page.  Doing this
for each output page of KMV pairs could be prohibitively expensive.  A
related issue is that, as described in the previous section, a hash
table is needed to match new keys with previously encountered keys.
But there is no guarantee that the hash table itself will not grow
arbitrarily large for data sets with many unique keys.  We need an
algorithm for generating KMV pairs that operates within a small
number of memory pages and performs a minimal number of passes through
the KV disk file.  An algorithm that meets these goals is diagrammed
in Figure \ref{fig:collate}; it reads the KV pages at most four times, and
writes out new KV or KMV pages at most three times.  It also uses a
finite-size in-memory hash table, that stores unique keys as they are
encountered while looping over the KV pages, as well as auxiliary
information needed to construct the output pages of KMV pairs.

@@

\item (Pass 1) The KV pairs are read, one page at a time, and split
into ``partition'' files, represented by KVp in the figure.  Each
partition file contains KV pairs whose unique keys (likely) fit
in the hash table (HT).  Initially, all KV pairs are assigned to the
first partition file as the HT is populated.  When the HT becomes
full, e.g. at the point represented by the horizontal line shown on
the leftmost vertical line as a downward scan is performed, the
fraction of KV pairs read thus far is used to estimate the number of
additional partition files needed.  As subsequent KV pairs are read,
they are assigned to the original partition file if the KV pair's
key is already in the current HT.  If not, a subset of bits in
the hash value of the key is used to assign the KV pair to one of the
new partition files.  The number of needed partition files is
estimated conservatively, rounding up to the next power-of-two, so
that extra passes through the KV pairs are almost never needed, and so
that the bit masking can be done quickly.  This first pass entails
both a read and write of all the KV pairs.

@@

\item (Pass 3) The unique keys in the HT for one partition are
scanned to determine the total size of KMV for keys in the HT.  
The associated values create KMV pairs that span
$M$ memory pages.  If $M > 1$, the associated partition file of KV
pairs is read again, and each KV pair is assigned to one of $M$
smaller ``set'' files, represented by KVs in the figure.  Each set
file contains the KV pairs that contribute to the KMV pairs that
populate one page of KMV output.  If a single KMV pair spans
multiple pages because it contains a very large number of values,
the corresponding KV pairs are still written to a single set
file.  Eventually, this pass reads all partition files, entailing both
a read and write of all KV pairs.

@@

The memory page requirements for the above algorithm are as follows.
Two contiguous pages of memory are used for the hash table.
Intermediate passes 2 and 3 can require numerous partition or set
files to be opened simultaneously and written to.  To avoid small
writes of individual KV datums, a buffer of at least 16K bytes is
allocated for each file.  The precise number of simultaneously open
files is difficult to bound, but typically one or two additional pages of 
memory suffice for all needed buffers.  Thus the total number of pages required
for the data reorganization stage of the {\it collate()} operation is
less than the seven used by the all-to-all communication stage.

Other MR-MPI library calls discussed in the previous section, such as
{\it clone()}, {\it collapse()}, and {\it gather()},
can also operate in out-of-core mode, typically with one pass
through the KV or KMV data and the use of one or two memory pages.
Sorting KV pairs, by key or value, is an exception.  An out-of-core
merge sort is performed as follows.  Two pages of KV datums are read
from disk.  Each is sorted using the C-library quicksort() function,
as discussed in the previous section.  The two pages are then scanned
and merged into a new file which is written to disk as its associated
memory page fills up.  This process requires five memory pages, which
includes vectors of KV datum pointers and lengths needed by the
in-memory quicksort() operation.  Once all pairs of pages have been
merged, the sort continues in a recursive fashion, merging pairs of
files into a new third file, without the need to perform additional
in-memory sorts.  Thus the overall memory requirement is five pages.  The
number of read/write passes through the KV data set for the merge sort
is $O(\log{M})$, where $M$ is the number of pages of KV pairs.
